---
title: "ENV 710 Initial Figures"
author: "S. Slay, E. Miles-Flynn, & Y.E. Su"
date: "2026-03-19"
output: pdf_document

---
```{r setup, include = TRUE, message = FALSE, warning = FALSE}

# Load the packages
library(here)
library(tidyverse)
library(moments)
library(dplyr)
library(viridis)
library(ggplot2)
library(ggpath)
library(gt)
library(paletteer)
library(grid)
library(gridExtra)
library(scales)
library(kableExtra)
library(knitr)
library(ggcorrplot)
library(readr)

```

```{r global_options, include = TRUE}

# Set global options for R chunks
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)

```

## Project files

The project is stored in the following Git repository: https://github.com/emilesflynn/ENV_710_Final_Project/

The raw data was downloaded from the Durham water quality web portal, which is managed by the City of Durham: http://www.durhamwaterquality.org/

## Data Wrangling

The raw data frame presents each measurement in a separate row. The code below merge the measurements conducted at the same location on the same day into one observation.

```{r raw data}
# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)

# Load your datasets
durham_data_raw <- read_csv(here("Data/SlaySMilesFlynnESuEData_Raw.csv"), 
    col_types = cols(Date = col_date(format = "%Y/%m/%d")))
# View(durham_data_raw)

# Pivot table and merge the measurements
df_wide <- durham_data_raw %>%
  pivot_wider(
    id_cols = c(`Station Name`, Date), 
    names_from = Parameter, 
    values_from = Value,
    values_fn = mean 
  )

write.csv(df_wide, file = here("Data/SlaySMilesFlynnESuEData_Cleaned.csv"), row.names = FALSE)
```

After obtaining the clean data frame, we then went on to add more explanatory or informational columns to help with further analyses. We isolated the river name and the sampling locations' distances to the river mouth, and add the year and month columns to help visualize the trends on the temporal scale.

```{r data_wrangling}
# Load the datasets
here()
data_clean <- read_csv(here("Initial Figures/SlaySMilesFlynnESuEData_Cleaned.csv"))

data_clean <- data_clean %>%
  mutate(
    #Extract the river name abbreviation from the station name
    River_abbr = str_extract(`Station Name`, "^[A-Z]{2}"),
    #Extract the distance to stream mouth
    Distance_miles = as.numeric(str_extract(`Station Name`, "[0-9.]+"))
  )

# Replace the abbreviations with your actual names
river_lookup <- data.frame(
  River_abbr = c("EL", "EN", "NH", "TF", "CC", "NE", "LC", "LL", "LR", "PN", "SI"),
  Watershed = c("Ellerbe Creek", "Eno River", "New Hope Creek", "Third Fork Creek", 
                      "Crooked Creek", "Northeast Creek", "Lick Creek", "Little Lick Creek", 
                      "Little River", "Panther Creek", "Stirrup Iron Creek")
)

# Merge the data frames
data_clean_final <- data_clean %>%
  left_join(river_lookup, by = "River_abbr")

# Add month and year columns
data_clean_final$Date <- ymd(data_clean_final$Date)

data_clean_final <- data_clean_final %>%
  mutate(Month = factor(month(Date),
                        levels = 1:12,
                        labels = month.abb,
                        ordered = TRUE),
         Year = year(Date),
         Storm_status = case_when(Date <= as.Date("2024-07-03") ~ "Before",
                            Date >= as.Date("2024-07-09") ~ "After"),
         Basin = case_when(
           River_abbr %in% c("EL", "EN", "LC", "LL", "LR", "PN", "SI") ~ "Neuse (North)",
           River_abbr %in% c("CC", "NE", "NH",  "TF") ~ "Cape Fear (South)"))

# Reorder columns
data_clean_final <- data_clean_final %>% 
  select(`Station Name`, Watershed, River_abbr, Distance_miles, Basin,
         Date, Month, Year, Storm_status, everything())

```

## Intial figures

### Scatterplot

```{r plot 1}
do_plot <- ggplot(data_clean_final, aes(x = Date, y = `Dissolved Oxygen`, color = Basin)) +
  # Add points and lines (grouped by Basin to show trends)
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE, method = "loess") + 
  
  # Add the NC DEQ Water Quality Standard line (5.0 mg/L)
  geom_hline(yintercept = 5.0, linetype = "dashed", color = "orange") +
  annotate("text", x = min(data_clean_final$Date), y = 5.3, 
           label = "NC Standard (5.0 mg/L)", color = "red", hjust = 0) +
  
  # Formatting
  scale_color_manual(values = c("Neuse (North)" = "#2c7bb6", 
                                "Cape Fear (South)" = "#d7191c")) +
  labs(title = "Dissolved Oxygen Levels in Durham Streams (2024)",
       subtitle = "Comparing Northern (Neuse) and Southern (Cape Fear) Watersheds",
       x = "Sampling Date",
       y = "Dissolved Oxygen (mg/L)") +
  theme_minimal() +
  theme(legend.position = "bottom")

do_plot
```
